---
title: "IS607 - Project 2 - Part 2"
author: "Rob Hodde"
date: "March 12, 2016"
output: html_document
---

Choose any three of the "wide" datasets identified in the Week 6 Discussion items.  Your code should be in an R Markdown file, posted to rpubs.com, and should include narrative descriptions of your data cleanup work, analysis, and conclusions.

---------

###Generator Capacity Prices###
***Problem Designed by Daniel Smilowitz***

---------

**The Data**

Generators in the NYS electricity market are paid based on the outcome of competitive Unforced Capacity (UCAP) auctions.

There are 3 auctions:

- *Monthly* - takes place monthly before final values are known

- *Spot* - takes place monthly once values are finalized

- *Strip* - takes place twice a year (May and November) for the following six months

The prices for each auction are set in four different locations - NYC, Long Island (LI), Lower Hudson Valley (LHV), and Rest of State (ROS). The past ~12.5 years worth of auction results are contained in a table, as depicted below.  All values are the price paid to generators in $/kW-month. 
Data source: New York Independent System Operator

---------

*Read the information from your .CSV file into R:*  

```{r}

suppressWarnings(library(knitr))
library(stringr)
suppressMessages(library(dplyr))
suppressWarnings(library(tidyr))

f <- read.csv(file = "https://raw.githubusercontent.com/LovinSpoonful/IS607-Project2/364b20a46f67b7b0742e9cb0b03853c51f97a791/UCAP.csv", header=FALSE, sep=",") #read the csv file into a data frame
kable(head(f), format = "markdown")

```


---------

*Use tidyr and dplyr as needed to tidy and transform your data.*

```{r, warning=FALSE}

typ <- as.vector(as.matrix(f[1,2:length(f[1,])]))  #get auction type
loc <- as.vector(as.matrix(f[2,2:length(f[2,])]))  #get auction location
typloc <- paste(typ,loc,sep = "|") #concatenate them temporarily so we can gather
names(f) <- c("AuctionDate",typloc)  # assign meaningful column names
f1 <- f[-(1:2),] # remove header rows

#create a single record for each auction (Date, Type and Location) 
f2 <- suppressWarnings(gather(f1,"TypLoc","Price",2:length(f1[1,])))  

#split the auction type and location into two columns
f3 <- separate(data = f2, col = "TypLoc", into = c("Type", "Location"), sep = "\\|")

#split the month and year into two columns
f4 <- separate(data = f3, col = "AuctionDate", into = c("Month", "Year"), sep = " ")

f4$Price <- str_trim(str_replace(f4$Price,"\\$","")) # remove dollar sign from prices
f4$Price <- as.numeric(as.character(f4$Price)) # convert prices to numeric values
f4 <- f4[complete.cases(f4),] # finally, remove all incomplete rows (have NA values)
```
```
Tidy Data Sample Output:
```
```{r, echo=FALSE}
kable(head(f4), format = "markdown") #now we have tidy data!

```

---------


**The Analysis**

*Which month of the year sees the highest prices in each location?*

```{r, warning=FALSE, message=FALSE}

#calculate the average price by month for each location
f6 <- f4 %>% 
  group_by(Location,Month) %>% 
  summarise(PriceAvg = mean(Price)) %>%  
  arrange(Location,desc(PriceAvg)) 

#select the highest average priced month for each location
f7 <- f6 %>%
  group_by(Location) %>%
  select(Month, PriceAvg) %>%
  top_n(1) %>%
  arrange(Location, desc(PriceAvg))
```
```
The highest prices by location occur as follows: 
```
```{r, echo=FALSE}
kable(f7, format = "markdown")
```

---------

*What is the average difference between NYC and ROS prices?*

```{r}
nyc <- subset(f7$PriceAvg, f7$Location == "NYC")
ros <- subset(f7$PriceAvg, f7$Location == "ROS")
```
```
The average difference between NYC and ROS 
prices is `r sprintf("$ %3.2f", nyc)` - `r sprintf("$ %3.2f", ros)` = `r sprintf("$ %3.2f", nyc - ros)`.
```
---------

*Which calendar year saw the highest average price across regions (ignoring weighting)?*

```{r, warning=FALSE, message=FALSE}
#calculate the average price for each year
f8 <- f4 %>% 
  group_by(Year) %>% 
  summarise(PriceAvg = mean(Price)) %>%  
  top_n(1) %>%
  arrange(desc(PriceAvg))
```
```
The highest average price across all regions occurred in `r f8$Year`.
```

---------

*Is the monthly auction or the spot auction more volatile (i.e. which has the most variability in pricing)?*
 
```{r}
monthlySD <- round(sd(subset(f4$Price,f4$Type == "Spot")),digits = 2)
spotSD    <- round(sd(subset(f4$Price,f4$Type == "Monthly")),digits = 2)
```
```
One measure of volatility is standard deviation. 
The standard deviation of the monthly auction is `r monthlySD`.
The standard deviation of the spot auction is `r spotSD`. 
Therefore the `r ifelse(spotSD > monthlySD,"spot","monthly")` auction is more volatile.

```

---------


